A statistically consistent approach to chemical tagging

### 2018-12-11 article/ms.tex
update ms.tex

\begin{abstract}
\end{abstract}

@@

\begin{itemize}
	\item Chemical tagging
	\item Limits on the efficacy of chemical tagging
	\item Clustering vs common traits
\end{itemize}

@@

% TODO: Is D a diagonal vector, or a matrix?! Make it consistent.

% Describe how we will attempt to define a generative model for the data, from the latent space.

In this work the data $\vecdata$ is a $\NumData \times \NumDimensions$ matrix 
where $\NumData$ is the number of stars and $\NumDimensions$ is the number of
chemical abundances measured for each star. We assume a
generative model for the data 
\begin{equation}
	\vecdata = \factorloads\factorscores + \vec{e}
	\label{eq:generative-model}
\end{equation}

\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ matrix of factor
loads that is common to all data points, and the factor scores for
the $\numdata$th data point
\begin{equation}
	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
\end{equation}
\noindent{}are drawn from the $\numcomponents$th multivariate  normal distribution.
The factor scores for all data points $\factorscores$ is then a 
$\NumData \times \NumLatentFactors$ matrix.
 We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \eye\specificvariance\right)$
is independent of the latent space, and $\specificvariance$ is a
diagonal matrix of $\NumDimensions$ entries. In Appendix~\ref{app:symbols}
we list all mathematical symbols used in this work and their description.

In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$. 
We assume that the latent space is lower dimensionality than the
data space (e.g., $\NumLatentFactors < \NumDimensions$).
Within the context of stellar abundances, the factor loads
$\factorloads$ can be thought of as the \emph{mean} yields
of nucleosynthetic
events (e.g., $s$-process yields from AGB stars), and the
factor scores are analogous to the relative counts of those 
nucleosynthetic events. The clustering in factor scores
achieves the same as a clustering procedure in data space,
except we simultaneously estimate the factor loads
(nucleosynthetic yields) that are common to all data points 
(stars). Within this framework a rare nucleosynthetic event
can still be described as a `common factor load' $\factorloads_\numlatentfactors$, 
but its rarity would be represented by associated factor
scores being zero for most stars and thus have no contribution
to the observed abundances. The factor loads and factor scores 
cannot be expressly interpreted like this because they have no
physical meaning, but this description of relative yields and 
rates should help build intuition for the model parameters,
and provide an astrophysical context.

One benefit of using a mixture of common factor analysers is
for modelling stellar chemical abundances is the scaling with
computational cost. If we considered data sets of order $3\times10^7$
entries (e.g., 30 chemical abundances for $10^6$ stars) purely as a
clustering problem, then even the most efficient clustering
algorithms would incur a significant cumulative computational 
overhead by searching the parameter space for the number of
clusters, and the optimal model parameters given that number
of components. However, because the mixture of factor analyzers
approach assumes that there is a \emph{lower dimensional latent 
space} in which the data are clustered, and that clustering is 
projected into real space by common factor loads, the 
dimensionality of the clustering problem is reduced from 
$N \times D$ to $N \times J$. This reduces computational cost through
faster execution of each E-M step, and on average fewer E-M steps
needed to reach a specified convergence threshold.

From a statistical standpoint, the primary advantage to using
a mixture of factor analysers is that we can simultaneously
estimate latent factors (e.g., infer nucleosynthetic 
yields) and perform clustering (e.g., chemical tagging) 
within a statistically consistent framework. That is to say
that we have a generative model for the data that includes
analogues of nucleosynthetic yields and star formation,
and the parameters of this model can be estimated consistently
given some data.

@@

%The model described by Equation~\ref{eq:generative-model} is indeterminate:
%there is no unique solution for the factor loads $\factorloads$ and scores
%$\factorscores$. However, an accurate estimate of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization algorithm \citep{EM}.

@@

The code associated with this \emph{Article} allows the model parameters
to be initialised using a variety of methods. For example, the initial factor 
loads $\factorloads$ can be drawn by one of the following methods:
\begin{enumerate}
	\item \emph{uniform}: a uniform distribution $\factorloads \sim \mathcal{U}\left(-1, +1\right)$,
	\item \emph{noise}: from noise $\factorloads \sim \mathcal{N}\left(0, 10^{-2}\right)$,
	\item \emph{SVD}: using the $\NumLatentFactors$ principal eigenvectors estimated by singular value decomposition,
	\item \emph{astrophysical}: using domain knowledge to set initial factor loads to vectors that are analogous to nucleosynthetic yields (see Section~\todo{\ref{sec:somewhere}}).
\end{enumerate}

We initially assign each data point as belonging to one of the $\NumComponents$
components either by random assignment (e.g., drawing from a multinomial distribution), 
or using the \texttt{K-means++} algorithm \citep{LLoyd} in latent space $\vecdata\factorloads$.
Given the initial factor loads and initial assignments, we then estimate the relative weights
$\vec\pi$, the mean factor scores of each component $\scoremeans$, and the covariance matrix
of factor scores of each component $\scorecovs$. In all scenarios we initialise the 
variance $\specificvariance$ in each dimension as 1.

\begin{comment}
We use the K-means++ algorithm to initially assign data points
randomly to $K$ clusters (in data space). This provides a so-called
hard initial association where each data point is wholly assigned
to a single component. Specifically we calculate the $n \times K$
responsibility matrix $\vec\tau$, where a entry of 1 indicates the
$n$th data point is associated with the $k$th cluster, and an
entry of 0 indicates no association such that
\begin{equation}
	\weight_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk} \quad .
\end{equation}
We then chose 1,000
random data points and factorize that matrix using singular-value
decomposition. We take the principal $p$ eigenvectors as the 
initial guess for the factor loads $\factorloads$, and calculate
an initial estimate of the factor score means
\begin{equation}
	\scoremeans_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk}\vecdata_{n}\factorloads
\end{equation}
\noindent{}and the covariance matrices of the clusters in latent space
\begin{equation}
	\scorecovs_k = \textrm{cov}{\left(\tau_{nk}\vecdata_{n}\right)} \quad .
\end{equation}
\end{comment}

@@

We use the expectation-maximization algorithm to maximise the log-likelihood
of the data, given the current estimate of the model parameters. With each step
cycle we update the model parameters and increase the log-likelihood.
At the expectation step we evaluate the log likelihood
given the model parameters $\vec\Psi$, and we re-calculate the responsibility
matrix whose entries are the posterior probability
that the $\numdata$th data point is associated to the $\numcomponents$th component,
given the data $\vecdata$ and the current estimate of the 
parameter vector $\vec\Psi$:
\begin{equation}
	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \eye\specificvariance)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\specificvariance)} \quad .
\end{equation}

% The complexity of SVD is O(n^3), and
% the complexity of K-means++ is O(log(k)). If we knew the complexity of
% k-means++ relative to N, then we would be able to make better decisions
% as to how to initialise: SVD first and then K-means in latent space?

% Discuss connection to PCA, FA, MFA.

% Include macros for \vec, \data, etc.
At the maximization step we update our estimates of the parameters,
conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
Initially we update the relative weights $\vec\weight\nextstep$ given
the responsibility matrix $\vec\tau$
\begin{equation}
	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents} \quad .
	% Include proof.
\end{equation}

@@

The updated estimates of the mean factor scores 
$\scoremeans\nextstep$ are given by
\begin{eqnarray}
	\scoremeans_\numcomponents\nextstep = \scoremeans_\numcomponents\thisstep + \frac{\vec{G}\transpose(\vecdata\transpose - \factorloads\thisstep\scoremeans_\numcomponents\thisstep)\vec\tau_\numcomponents}{\NumData\weight_\numcomponents\nextstep}
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
	\vec{W} &=& (\scorecovs_\numcomponents\thisstep)^{-1}\eye \\
	\vec{V} &=& \left(\specificvariance\thisstep\right)^{-1} \\
	\vec{C} &=& (\vec{W} + (\factorloads\thisstep)\transpose\vec{V}\factorloads\thisstep)^{-1}\eye \\
	\vec{G} &=& \left[\vec{V} - \vec{V}\factorloads\thisstep\vec{C}\left(\vec{V}\factorloads\thisstep\right)\transpose\right]\factorloads\thisstep\scorecovs_k\thisstep \quad .
\end{eqnarray}

@@

\noindent{}where $\odot$ denotes is the entrywise (Hadamard) product.

@@

\begin{comment}
For this subset
we examined the measured abundances and
counted the number of stars with \texttt{flag\_x\_fe = 0}
for each abundance. This flag indicates that there is no known
reason to suspect that the measured chemical abundance is
unreliable.

We sorted the elemental abundances by the number of stars
with reliable measurements, from highest to lowest. Starting
with Fe, we iterated over each element and counted the number
of stars with reliable measurements of that element and all
other elements earlier in the list. We illustrate this process
in Figure~\ref{fig:galah-abundances}, where we show that
if we include the top 14 elemental abundances, we would have
12,798 stars with reliable measurements in all fourteen elements.
This is relevant to the approach we use here, as the method
described here requires no missing data.
\end{comment}

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/galah-abundances.png}
    \caption{Number of \Galah\ stars with reliable abundance measurements,
    		 shown in order from the most commonly measured elemental abundance
		 	 (Fe) to the least reliable element (Li). The solid line
			 shows the number of reliable measurements for the given element,
			 and the dotted (blue) line indicates the stars with reliable
			 abundances in all elements listed to the left (including the
			 element of interest). For example, there are $\sim60,000$
			 stars with reliable abundances in all elements listed from
			 Fe to Cu, inclusive.}
    \label{fig:galah-abundances}
\end{figure}

\begin{comment}
This leaves us with a sample of 12,798 stars each with
reliable stellar parameters ($\teff$, $\logg$) and fourteen
(reliably) measured abundances: Na, Mg, Si, Ca, Sc, Ti, Cr, Mn, 
Fe, Ni, Cu, Zn, Ba, and Eu
\todo{Y} (reliable) measured abundances: \todo{Mn, Fe}.
These elements are produced from multiple nucleosynthetic
pathways, including \todo{X}. 
It is this subsample that we will use for experiments
in chemical tagging.
\end{comment}

>>>>

\begin{abstract}
Chemical tagging promises to distinguish unique star formation sites based on the
present day photospheric abundances of stars.  It is not known how effective this
will be.  Clustering techniques used for chemical tagging  assume statistical 
independence between abundances. Theory and observations agree that this assumption
is incorrect: some elemental abundances are produced through multiple nucleosynthetic
processes, and most nucleosynthetic processes generates multiple chemical abundances.
Here we introduce a statistically consistent approach to chemical tagging, where we
simultaneously infer the nucleosynthetic processes that produce chemical abundances,
and we perform clustering in this lower dimensional latent space. \todo{we use galah}
\todo{scales iwth data}
\end{abstract}

@@

\begin{itemize}
	\item Chemical tagging
	\item Limits on the efficacy of chemical tagging
	\item Clustering vs common traits
	\item Dimensionality of data space
\end{itemize}

@@

%Factor analysis is a statistical method used to describe variability among observed, correlated variables in terms of a potentially lower number of unobserved variables called factors.

Factor analysis is a common statistical approach for describing correlated 
observations with a lower number of latent variables \citep{factor-analysis}.
\todo{Analagous approaches used in astrophysics: pca, etc}.

Similarly, clustering techniques are commonly used to identify groups in 
data space.

Here we use a combination of these methods that provides a
consistent description of the data, given the latent variables and their 
clustering in latent space. In this work the data $\vecdata$ is a 
$\NumData \times \NumDimensions$ matrix where $\NumData$ is the number of 
stars and $\NumDimensions$ is the number of chemical abundances measured 
for each star. We assume a
generative model for the data 
\begin{equation}
	\vecdata = \factorloads\factorscores + \vec{e}
	\label{eq:generative-model}
\end{equation}

\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ matrix of factor
loads that is common to all data points, and the factor scores for
the $\numdata$th data point
\begin{equation}
	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
\end{equation}
\noindent{}are drawn from the $\numcomponents$th multivariate  normal distribution.
The factor scores for all data points $\factorscores$ is then a 
$\NumData \times \NumLatentFactors$ matrix.
 We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \eye\specificvariance\right)$
is independent of the latent space, and $\specificvariance$ is a
diagonal matrix of $\NumDimensions$ entries. In Appendix~\ref{app:symbols}
we list all mathematical symbols used in this work and their description.
In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$.

We assume that the latent space is lower dimensionality than the
data space (e.g., $\NumLatentFactors < \NumDimensions$).
Within the context of stellar abundances, the factor loads
$\factorloads$ can be thought of as the \emph{typical} yields
of nucleosynthetic
events (e.g., $s$-process production from AGB stars), and the
factor scores are analogous to the relative counts of those 
nucleosynthetic events. The clustering in factor scores
achieves the same as a clustering procedure in data space,
except we simultaneously estimate the latent processes that are
common to all stars (the so-called factor loads, analogous to 
nucleosynthetic yields). Within this framework a rare nucleosynthetic event
can still be described as a `factor load' $\factorloads_\numlatentfactors$, 
but its rarity would be represented by associated factor
scores being zero for most stars and thus have negligible contribution
to the observed abundances. In practice the factor loads can only be 
identified up to orthogonality and cannot be expressly interpreted as
nucleosynthetic yields because they have limited physical meaning,
but this description of typical yields and relative event rates should
help build intuition for the model parameters, and provide context
within the astrophysical problem it is being applied.

\todo{this model of orthogonality takes the independence into account}

Another benefit of using a mixture of common factor analysers is
for modelling stellar chemical abundances is the scaling with
computational cost. If we considered data sets of order $3\times10^7$
entries (e.g., 30 chemical abundances for $10^6$ stars) purely as a
clustering problem, then even the most efficient clustering
algorithms would incur a significant cumulative computational 
overhead by searching the parameter space for the number of
clusters, and the optimal model parameters given that number
of components. However, because the mixture of factor analyzers
approach assumes that there is a \emph{lower dimensional latent 
space} in which the data are clustered, and that clustering is 
projected into real space by common factor loads, the 
dimensionality of the clustering problem is reduced from 
$N \times D$ to $N \times J$. This reduces computational cost through
faster execution of each E-M step, and on average fewer E-M steps
needed to reach a specified convergence threshold.

From a statistical standpoint, the primary advantage to using
a mixture of factor analysers is that we can simultaneously
estimate latent factors (e.g., infer nucleosynthetic 
yields) and perform clustering (e.g., chemical tagging) 
within a statistically consistent framework. That is to say
that we have a generative model for the data that can 
quantitatively account for nucleosynthetic yields, variations in
turbulence and gas mixing, or star formation efficiency,
%can account
%analogues of nucleosynthetic yields and star formation,
and the parameters of this model can be estimated consistently
given some data.

@@

As mentioned previously, the model as described is indeterminate in that
there is no unique solution for the factor loads $\factorloads$ and scores
$\factorscores$. These quantities can only be determined up until 
orthogonality. However, accurate estimates of the parameter vector $\vec\Psi$
can be obtained by the expectation-maximization algorithm \citep{EM}.
%The model described by Equation~\ref{eq:generative-model} is indeterminate:
%there is no unique solution for the factor loads $\factorloads$ and scores
%$\factorscores$. However, an accurate estimate of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization algorithm \citep{EM}.

@@

Here we describe the default initialisation of the model parameters, which
we have found to be robust in a large set of models. Given a number of
$\NumLatentFactors$ latent factors we initialise the factor load
$\factorloads$ entries to be drawn from a uniform distribution
$\factorloads \sim \mathcal{U}\left(-1, +1\right)$ and then ensure that
the factor loads are orthogonal such that
%TODO: Should we re-write this so that we just draw from the scipy special ortho group?
\begin{equation}
	\factorloads \factorloads\transpose = \eye \quad .
\end{equation}
%        AL = linalg.cholesky(A.T @ A)
%        A = A @ linalg.solve(AL, np.eye(self.n_latent_factors))
% 	
We then initially assign each data point as belonging to one of the
$\NumComponents$ components by drawing from a multinomial distribution
and generating the responsibility matrix $\vec\tau$. Given the initial
factor loads and assignments, we then estimate the relative weights
$\vec\pi$, the mean factor scores of each component $\scoremeans$, and
the covariance matrix of factor scores of each component $\scorecovs$.
Finally, we initialise the specific variance $\specificvariance$ in each
dimension as 1.

Other initialisation methods are available in the code 
associated with this \emph{Article}. In later Sections we show that 
the optimised model parameters are insensitive to the choice initialisation, 
up until orthogonal rotation of the latent space.

@@

We use the expectation-maximization algorithm to estimate the model parameters.
With each expectation step we evaluate the log likelihood given the model 
parameters $\vec\Psi$, and we re-calculate the responsibility 
matrix $\vec\tau$ whose entries are the posterior probability
that the $\numdata$th data point is associated to the $\numcomponents$th component,
given the data $\vecdata$ and the current estimate of the 
parameter vector $\vec\Psi$:
\begin{equation}
	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \eye\specificvariance)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\specificvariance)} \quad .
\end{equation}

At the maximization step we update our estimates of the parameters,
conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
First we estimate the relative weights $\vec\weight\nextstep$ given
the responsibility matrix $\vec\tau$
\begin{equation}
	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents}
	% Include proof.
\end{equation}
\noindent{}where the $\vec{X}\thisstep$ superscript refers to the current estimate of a variable
and $\vec{X}\nextstep$ refers to the updated estimate for the next iteration.
%\footnote{Here $\vec\textrm{X}\now$ refers to the current estimate of a variable $\vec\textrm{X}$ and $\vec\textrm{X}\nextstep$ refers to the updated estimate for the next iteration.}

@@

The updated estimates of the mean factor scores 
$\scoremeans\nextstep$ for each component are then given by
\begin{eqnarray}
	\scoremeans_\numcomponents\nextstep = \scoremeans_\numcomponents\thisstep + \frac{\vec{G}\transpose(\vecdata\transpose - \factorloads\thisstep\scoremeans_\numcomponents\thisstep)\vec\tau_\numcomponents}{\NumData\weight_\numcomponents\nextstep}
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
	\vec{W} &=& (\scorecovs_\numcomponents\thisstep)^{-1}\eye \\
	\vec{V} &=& \left(\specificvariance\thisstep\right)^{-1} \\
	\vec{C} &=& (\vec{W} + (\factorloads\thisstep)\transpose\vec{V}\factorloads\thisstep)^{-1}\eye \\
	\vec{G} &=& \left[\vec{V} - \vec{V}\factorloads\thisstep\vec{C}\left(\vec{V}\factorloads\thisstep\right)\transpose\right]\factorloads\thisstep\scorecovs_k\thisstep \quad .
\end{eqnarray}

@@

\noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. We repeated
the expectation-maximization algorithm for up to 1,000 steps or until the log-likelihood
improved by less than $10^{-5}$ between successive iterations. Although our implementation
allows for a positive regularisation term to be added along the diagonal of the covariance
matrices in latent space $\scorecovs$, we set this regularisation term to zero for all
experiments performed here.

@@
